---
title: "AlberdiLab | Río-López et al. 2026"
subtitle: "Study in progress"
author:
  - Raquel Río-Lopez, Antton Alberdi
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/pig_stress_hologenomics
description: |
  Stress pig hologenomics project
link-citations: yes
github-repo: alberdilab/pig_stress_hologenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across ponds included in a restoration plan.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/pig_stress_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)
library(janitor)
library(broom)

library(purrr)
library(dplyr)
library(tibble)
library(ggplot2)

if(!requireNamespace("BiocManager", quietly=TRUE))
 + install.packages("BiocManager")
library(BiocManager)
BiocManager::install("SummarizedExperiment")
library(SummarizedExperiment)
BiocManager::install("edgeR")
library(edgeR)
BiocManager::install("limma")
library(limma)

```

#Differential Expression Analyses

##Pre-processing of data

###Adapt the format

Define a function that reads and processes RNASeq input files, adapting the format to produce matrices containing only integer count values, with genes as rows and samples as columns (without annotation columns).


```{r read_and_adapt_function, warning=FALSE, comments="", message=FALSE}

read_and_adapt <- function(file_path) {
  counts <- read_tsv(file_path, col_names = TRUE)
  counts_matrix <- counts %>%
    select(-gene_name) %>%
    column_to_rownames(var = "gene_id") %>%
    as.matrix()
  mode(counts_matrix) <- "numeric"
  return(counts_matrix)
}
```
Apply the function to all the tissues (amygdala, cortex, hyppocampus and liver).

```{r read_and_adapt_input_files, warning=FALSE, comments="", message=FALSE}
amygdala_counts <- read_and_adapt("RNASeq_tables/RNASeq_out_Amigdala/salmon.merged.gene_counts.tsv")
cortex_counts <- read_and_adapt("RNASeq_tables/RNASeq_out_Cortex/salmon.merged.gene_counts.tsv")
hippocampus_counts <- read_and_adapt("RNASeq_tables/RNASeq_out_Hippocampus/salmon.merged.gene_counts.tsv")
liver_counts <- read_and_adapt("RNASeq_tables/RNASeq_out_Liver/salmon.merged.gene_counts.tsv")
head(amygdala_counts)
```

###Filter low-expressed genes

Keep genes with counts per million (CPM) values greater than one in at least two samples. The counts per million (CPM) value is used instead of the read count to eliminate the deviation caused by different sequencing depths.

```{r filter_function, warning=FALSE, comments="", message=FALSE}

filter_CPM <- function(counts_matrix, min_cpm = 1, min_samples = 2){
  cpm_values <- cpm(counts_matrix)
  keep <- rowSums(cpm_values > min_cpm) >= min_samples
  filtered_matrix <- counts_matrix[keep, , drop=FALSE]
  return(filtered_matrix)
}

```
Apply to all tissues.

```{r filter_tissue_matrices, warning=FALSE, comments="", message=FALSE}
amygdala_counts_f <- filter_CPM(amygdala_counts)
cortex_counts_f   <- filter_CPM(cortex_counts)
hippocampus_counts_f <- filter_CPM(hippocampus_counts)
liver_counts_f    <- filter_CPM(liver_counts)
```

##Import sample metadata

Import all the sample information regarding treatment condition, sex and day of sacrifice for all tissues.

```{r all_metadata, warning=FALSE, comments="", message=FALSE}

#Read metadata
metadata_amygdala <- read.table("metadata/metadata_amygdala.txt", header = TRUE)
metadata_cortex <- read.table("metadata/metadata_cortex.txt", header = TRUE)
metadata_hippocampus <- read.table("metadata/metadata_hippocampus.txt", header = TRUE)
metadata_liver <- read.table("metadata/metadata_liver.txt", header = TRUE)

#Make sure sample order is the same in data and metadata
metadata_liver <- metadata_liver[
  match(colnames(liver_counts), metadata_liver$sample),
]

head(metadata_amygdala)
```


##Limma voom

First, we create a function to apply Limma voom analysis to all tissues. In this function, the first step is to create a design matrix specific for Limma voom analysis using a group vector. Then create a DGEList object and normalise the data by the trimmed mean of M-values (TMM) method and apply voom. 

```{r limma_voom_function, warning=FALSE, comments="", message=FALSE}

run_limma_voom <- function(counts_matrix, metadata, group_col = "condition", covariates = c("sex"), logFC_thresh = 2, pval_thresh = 0.05) {

  #Create group (factor) using condition column
  group <- factor(metadata[[group_col]], levels = c("control", "stress"))
  
  #Build the design formula including group and covariables
  design_formula <- as.formula(paste("~", group_col, "+", paste0(covariates, collapse = "+")))
  design <- model.matrix(design_formula, data = metadata)
  rownames(design) <- metadata$sample
  
  #Create DGEList and normalise by TMM method
  dge <- DGEList(counts = counts_matrix, group = group)
  dge <- calcNormFactors(dge, method = "TMM")
  
  #Apply voom
  v <- voom(dge, design, plot = FALSE)
  
  #Adjust the linear model
  fit <- lmFit(v, design)
  fit <- eBayes(fit)
  
  #Extract result table
  res <- topTable(fit, coef = "conditionstress", n = Inf, sort.by = "none")
  
  #Create significance column based on thresholds
  res$sig <- factor(
    ifelse(res$adj.P.Val < pval_thresh & abs(res$logFC) > logFC_thresh,
           ifelse(res$logFC > logFC_thresh, "up", "down"),
           "not"),
    levels = c("up", "down", "not")
  )
  
  #Summary of DEGs
  deg_summary <- summary(res$sig)
  
  #Results
  return(list(results = res, dge = dge, v = v, design = design, deg_summary = deg_summary))
}

```

Apply the function to all the tissues.

```{r limma_voom_tissues, warning=FALSE, comments="", message=FALSE}

#Group data and metadata into lists
counts_list <- list(amygdala = amygdala_counts_f, cortex = cortex_counts_f, hippocampus = hippocampus_counts_f, liver = liver_counts_f)
metadata_list <-list(amygdala = metadata_amygdala, cortex = metadata_cortex, hippocampus = metadata_hippocampus, liver = metadata_hippocampus)

#Apply the function to each pair
limma_results <- map2(counts_list, metadata_list, ~ run_limma_voom(counts_matrix = .x, metadata = .y))

#DEGs summary for each tissue
map(limma_results, "deg_summary")

#Save result tables
walk2(limma_results, names(limma_results), ~ write.csv(.x$results, file = file.path("differential_analysis/limma_voom", paste0("result_limma_voom_", .y, ".csv")), row.names = TRUE))

```

Create a volcano plot function.

```{r limma_voom_volcano, warning=FALSE, comments="", message=FALSE}

# Volcano plot function
volcano_plot <- function(data, thre_logFC = 2, thre_pval = 0.05) {
  data$color <- as.character(
    ifelse(data$adj.P.Val < thre_pval & abs(data$logFC) > thre_logFC,
           ifelse(data$logFC > thre_logFC, 'red', 'green'),
           'grey'))
  
  p <- ggplot(data, aes(x = logFC, y = -log10(adj.P.Val))) +
    geom_point(size = 0.25, colour = data$color) +
    theme_bw() +
    geom_hline(yintercept = -log10(thre_pval), colour = "black", linetype = "dashed") +
    geom_vline(xintercept = -thre_logFC, colour = "black", linetype = "dashed") +
    geom_vline(xintercept = thre_logFC, colour = "black", linetype = "dashed") +
    theme(axis.text.y = element_text(size = 10, face = "plain"),
          axis.title.y = element_text(size = 10, face = "bold"),
          panel.border = element_blank(),
          axis.line = element_line(colour = "black", size = 0.8),
          legend.position = "none",
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_rect(fill = "white")) +
    ylab("-log10(adj.P.Val)") +
    xlab("log2FC")
  
  return(p)
}

```
Apply the function to all the tissues and save the plots.


```{r limma_voom_volcano_tissues, warning=FALSE, comments="", message=FALSE}
#Define output directory
output_dir <- "differential_analysis/limma_voom"
volcano_dir <- file.path(output_dir, "volcano_plots")
if (!dir.exists(volcano_dir)) dir.create(volcano_dir, recursive = TRUE)

# Iterate over the results and save each volcano plot
walk2(limma_results, names(limma_results), function(res, tissue) {
  # Extract results
  df <- res$results
  # Generate plot
  p <- volcano_plot(df, thre_logFC = 2, thre_pval = 0.05)
  # Save as png
  ggsave(filename = file.path(volcano_dir, paste0("volcano_", tissue, ".png")),
         plot = p, width = 6, height = 5, dpi = 300)
  # Save as pdf
  ggsave(filename = file.path(volcano_dir, paste0("volcano_", tissue, ".pdf")),
         plot = p, width = 6, height = 5)
})

for (tissue in names(limma_results)) {
  df <- limma_results[[tissue]]$results
  p <- volcano_plot(df, thre_logFC = 2, thre_pval = 0.05)
  print(p)
  cat("\n\n**Volcano plot for", tissue, "tissue**\n\n")
}

```


##EdgeR

##DESeq2

