---
title: "AlberdiLab | Río-López et al. 2026"
subtitle: "Study in progress"
author:
  - Raquel Río-Lopez, Antton Alberdi
date: "Last update: `r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/pig_stress_hologenomics
description: |
  Stress pig hologenomics project
link-citations: yes
github-repo: alberdilab/pig_stress_hologenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of gut microbiomes of newts across ponds included in a restoration plan.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/pig_stress_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)
library(janitor)
library(broom)
```
